String Constraints with Regex-Counting and String-Length Solved Efficiently

### 2023-05-04 2023-1-19/sections/algorithms.tex
from RECL to CEFA

At first, we show how to construct a CEFA from a regular expression where counting operators may occur.

Let $e$ be a regular expression and $(e')^{\{m,n\}}$ be an occurrence of the counting operators in $e$. Then $(e')^{\{m,n\}}$ is said to be \emph{persistent} in $e$ if  $(e')^{\{m,n\}}$ is not in the scope of a complement, Kleene star, or counting operator in $e$.

@@

% We must syntactically rewrite bounded repetition $\regex\{m,n\}$ if it is the sub-regex of complement (e.g., $(\regex\{m,n\})^C$), closure(e.g., $(\regex\{m,n\})^*$), and bounded repetition(e.g., $(\regex\{m,n\})\{m', n'\}$), we unwind it to $\regex^m\mid\cdots\mid\regex^n$. After this syntactic rewriting, we call the resulting regex $\regex'$ \emph{non-nested}.
Given a regex $\regex$ without an unhandled counting operator, the base case is constructing the CEFA of a single character, an empty string, or an empty language. The inductive step is constructing CEFA of the concatenation, union, intersection, closure, complement, and counting of CEFA(s) of sub-regex(es). Operations except for counting are trivial and represented in Appendix , we only illustrate counting here.

>>>>

At first, we show how to construct a CEFA from a regular expression where counting operators may occur. 
Let us start with register-representable regular expressions defined in the sequel.

Let us fix an alphabet $\Sigma$.

Let $e$ be a regular expression over $\Sigma$. Then an occurrence of counting operators in $e$, say $(e')^{\{m,n\}}$ (or $(e')^{\{m, \infty\}}$), is said to be \emph{register-representable} if $(e')^{\{m,n\}}$ (or $(e')^{\{m, \infty\}}$) is not in the scope of a Kleene star, another counting operator, complement, or language difference in $e$. We say that $e$ is \emph{register-representable} if all occurrences of counting operators in $e$ are register-representable. For instance, $(a^{\{2, 6\}} \concat b^*) \cap (a + b)^{\{4, \infty\}}$ is register-representable, while $(a^{\{2, 6\}} \concat b^*) \setminus (a + b)^{\{4, \infty\}}$ and $(a^{\{2, 6\}} \concat b^*)^{\{1,\infty\}}$ are not.

Let $e$ be a register-representable regular expression over $\Sigma$. We are going to construct a CEFA out of $e$, denoted by $\aut_e$, by the following procedure. 
\begin{enumerate}
\item For each maximal sub-expression $e'$ of $e$ such that $e'$ contains no occurrences of counting operators, an NFA $\aut_{e'}$ can be constructed by a structural induction on the syntax of $e'$. 
%
\item For each sub-expression $(e')^{\{m,n\}}$ with $m \le n$ (resp. $(e')^{\{m, \infty\}}$) of $e$, we construct a CEFA $\aut_{(e')^{\{m,n\}}}$ (resp. $\aut_{(e')^{\{m, \infty\}}}$) as follows. Let $\aut_{e'} = (Q, \Sigma, \delta, I, F)$. From the construction for Kleene star in Section~\ref{sec:pre}, $(\aut_{e'})^* = (Q', \Sigma, \delta', I', F')$, where $Q' = Q \cup \{q_0\}$ with $q_0 \not \in Q$, $I' = \{q_0\}$, $F'= F \cup \{q_0\}$, $\delta' = \delta  \cup \{(q_0, a, q')  \mid \exists q'_0 \in I.\ (q'_0, a, q') \in \delta\} \cup \{(q, a, q') \mid q \in F, \exists q'_0 \in I.\ (q'_0, a, q') \in \delta\}$. Then $\aut_{(e')^{\{m,n\}}} = ((r'), Q', \Sigma, \delta'', I', F', \alpha')$, where $r'$ is a register, \\
%
$$
\begin{array}{l c l}
\delta'' & = & \{(q, a, q', (0)) \mid (q, a, q') \in \delta\}\  \cup \\
& & \{(q_0, a, q', (1))  \mid \exists q'_0 \in I.\ (q'_0, a, q') \in \delta\}\ \cup \\
& & \{(q, a, q', (1)) \mid q \in F, \exists q'_0 \in I.\ (q'_0, a, q') \in \delta\},
\end{array}
$$ 
%
moreover, $\alpha' = m \le r' \le n$ if $I \cap F = \emptyset$, and $\alpha' = r' \le n$ otherwise.  (Intuitively, if $\varepsilon$ is accepted by $\aut_{e'}$, then the value of $r'$ can be less than $m$.) Finally, $\aut_{(e')^{\{m, \infty\}}}$ is constructed by adapting $\alpha'$ in $\aut_{(e')^{\{m,n\}}}$ as follows: $\alpha' = m \le r'$ if $I \cap F = \emptyset$ and $\alpha' = \ltrue$ otherwise.

%
\item For each sub-expression $e'$ of $e$ such that $e'$ contains occurrences of counting operators but $e'$ itself is not of the form $(e'_1)^{\{m,n\}}$ or $(e'_1)^{\{m,\infty\}}$, from the assumption that $e$ is register-representable, we know that $e'$ is of the form $e'_1 \concat e'_2$, $e'_1 + e'_2$, $e'_1 \cap e'_2$, or $(e'_1)$. For $e' = (e'_1)$, we have $\aut_{e'} = \aut_{e'_1}$. For $e' = e'_1 \concat e'_2$, $e' = e'_1 + e'_2$, or $e' = e'_1 \cap e'_2$, suppose that CEFA $\aut_{e'_1}$ and $\aut_{e'_2}$ have been constructed. Without loss of generality,  we assume that $R_{\aut_{e'_1}} \cap R_{\aut_{e'_2}} = \emptyset$. Then we have $\aut_{e'} = \aut_{e'_1} \concat \aut_{e'_2}$, $\aut_{e'} = \aut_{e'_1} \cup \aut_{e'_2}$, $\aut_{e'} = \aut_{e'_1} \cap \aut_{e'_2}$ respectively. 
\end{enumerate}

\begin{example}
an example here
\end{example}

\zhilin{stopped here}

@@

% We must syntactically rewrite bounded repetition $\regex\{m,n\}$ if it is the sub-regex of complement (e.g., $(\regex\{m,n\})^C$), closure(e.g., $(\regex\{m,n\})^*$), and bounded repetition(e.g., $(\regex\{m,n\})\{m', n'\}$), we unwind it to $\regex^m\mid\cdots\mid\regex^n$. After this syntactic rewriting, we call the resulting regex $\regex'$ \emph{non-nested}.
Given a regex $\regex$ without an unhandled counting operator, the base case is constructing the CEFA of a single character, an empty string, or an empty language. The inductive step is constructing CEFA of the concatenation, union, intersection, closure, complement, and counting of CEFA(s) of sub-regex(es). Operations except for counting are trivial and represented in Appendix \ref{appendix:cefa}, we only illustrate counting here.
